---
title: "Bayesian Inference"
author: "Ivan Nikolov"
date: "17/5/2042"
output: pdf_document
---
lib
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r imports}
library(ggplot2)
library(rstanarm)
library(rstan)
library(bayesplot)
library(bayestestR)
#library(see)
#library(parameters)
set.seed(42)
```

## Including Plots



```{r read_data, echo=FALSE}
df <- read.csv('./data/dataset.csv', header = TRUE)
df
```

```{r standardize_data}
df$Angle <- (df$Angle - mean(df$Angle)) / sd(df$Angle)
df$Distance <- (df$Distance - mean(df$Distance)) / sd(df$Distance)
```

After standardizing the data, we like to choose priors for our model. For the distance, I my prior thinking is that a larger distance will likely negatively affect the shot made. For that I chose a normal distribution with a mean -1 and a stardard deviation 2.5 (the default from the package).

For the angle I also think that a larger angle will likely negatively affect the shot made, but to a lesset extent than the distance. For that I chose a normal distribution with a mean -0.5 and a stardard deviation 2.5 (the default from the package).

```{r priors}
priors <- list(
  normal(-1, 2.5),          # Prior for Angle coefficient
  normal(-0.5, 2.5)          # Prior for Distance coefficient with negative mean
)
```

```{r model}
priors <- normal(location=c(-1, -0.5),
                 scale= c(2.5, 2.5)
)


basket_lg <- rstanarm::stan_glm(
  Made ~ Angle + Distance,
  df,
  family = binomial(link = "logit"),
  prior = priors,
  iter = 10000,
  chain = 4
)
samples <- extract(basket_lg$stanfit)
paste("We have obtained: ", length(samples$alpha), " samples")
prior_summary(basket_lg) #priors used
parameters(basket_lg) #parameters summary
```

```{r posterior}
mcmc_dens(basket_lg)
```

```{r posterior_plots}
ggplot(data = data.frame(x=samples$beta[,2], y=samples$beta[,1]), aes(x = x, y=y)) +
  geom_point(size=0.1) +
  ggtitle('Scatter plot of values for betas')+
  xlab("Distance")+
  ylab("Angle")+
  coord_fixed()+
  xlim(-1,0)+
  ylim(-0.5,0.5)+
  geom_density2d() +
  geom_hline(yintercept = mean(samples$beta[,1]), linetype = "dashed", color = "red") +
  geom_vline(xintercept = mean(samples$beta[,2]), linetype = "dashed", color = "red")
  
```


```{r plots}
mean_x <- mean(samples$beta[,2])
mean_y <- mean(samples$beta[,1])

ggplot(data = data.frame(x=samples$beta[,2], y=samples$beta[,1]), aes(x = x, y=y)) +
  geom_point(size=0.1) +
  #ggtitle('Scatter plot of values for betas') +
  xlab("x - Distance") +
  ylab("y - Angle") +
  coord_fixed() +
  xlim(-1,0) +
  ylim(-0.5,0.5) +
  geom_density2d() +
  geom_hline(yintercept = mean_y, linetype = "dashed", color = "red") +
  geom_text(data = data.frame(x = mean_x, y = mean_y), aes(x = 0, y = y+0.04, label = paste("y =", round(y, 2))), 
            vjust = 0, hjust = 0.9, color = "red") +
  geom_vline(xintercept = mean_x, linetype = "dashed", color = "red") +
  geom_text(data = data.frame(x = mean_x, y = mean_y), aes(x = x+0.04, y = -0.5, label = paste("x =", round(x, 2))), 
            vjust = 0, hjust = 0, color = "red")
```
```{r Q1}
mean(abs(samples$beta[,2]) - abs(samples$beta[,1]) > 0.001)
```
```{r posteriors positive negative}
plot(p_direction(basket_lg))

```
```{r Q2}
mean(samples$beta[,1] < 0)
```
```{r Q2}
mean(samples$beta[,1] > 0)
```

